Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-11-19 ms/ms.Rmd
<edit index>

>>>>

Elliptical bifaces have a deep history in eastern North America extending from the Middle Archaic to well past European contact. In many instances they have been deposited in caches, which may represent large, corporate bundles decommissioned amid polity demise or radical transformations in social practice. Stylistic changes are evident in both length and width for these bifaces, and result in great variability in shape and size. While some undoubtedly were knapped for every-day, mundane uses, others were crafted from exotic cherts and reflect non-utilitarian uses. The _oversize_ or _exaggerated_ forms, mimicking combat weapons, are often referred to as symbolic weaponry. Such objects are typically those “whose size, delicacy, material of manufacture, and craftsmanship are outside the bounds of any weapon that might have been actually used” \citep[624]{RN11177}.

The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South \citep{RN11179}, are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” \citep[143]{RN11178}. In the western Middle Tennessee Valley, the Pickwick Burial Complex \citep[94]{RN11180}, dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken. Elliptical bifaces find their most dramatic expression as Caddoan and Mississippian exaggerated, symbolic weaponry. Mississippian Ramey knives, dating from c. CE 1000 to 1300 are found associated with Cahokia political dynamics. The slightly later Duck River sword-form bifaces (CE 1275-1400) continue the emphasis on high-quality knapping and long-distance circulation. Such lanceolate blades, although smaller, continue into the early seventeenth century when they serve as regalia among political and social elites in ritual performances.

### 2022-11-19 ms/ms.tex
<edit index>

>>>>

Elliptical bifaces have a deep history in eastern North America
extending from the Middle Archaic to well past European contact. In many
instances they have been deposited in caches, which may represent large,
corporate bundles decommissioned amid polity demise or radical
transformations in social practice. Stylistic changes are evident in
both length and width for these bifaces, and result in great variability
in shape and size. While some undoubtedly were knapped for every-day,
mundane uses, others were crafted from exotic cherts and reflect
non-utilitarian uses. The \emph{oversize} or \emph{exaggerated} forms,
mimicking combat weapons, are often referred to as symbolic weaponry.
Such objects are typically those ``whose size, delicacy, material of
manufacture, and craftsmanship are outside the bounds of any weapon that
might have been actually used'' \citep[624]{RN11177}.

The earliest elliptical bifaces, associated with Middle Archaic period,
recovered as Benton caches (4500---4000 BCE) in the Mid-South
\citep{RN11179}, are clearly non-utilitarian, as they ``meet the
expectations of sacred markers for secular exchange''
\citep[143]{RN11178}. In the western Middle Tennessee Valley, the
Pickwick Burial Complex \citep[94]{RN11180}, dating to the Late
Archaic/Gulf Formational period (3000---300 BCE) exhibits a continuation
of bipointed blades, some notched or truncated, found in ritual caches
and ceremonially broken. Elliptical bifaces find their most dramatic
expression as Caddoan and Mississippian exaggerated, symbolic weaponry.
Mississippian Ramey knives, dating from c.~CE 1000 to 1300 are found
associated with Cahokia political dynamics. The slightly later Duck
River sword-form bifaces (CE 1275-1400) continue the emphasis on
high-quality knapping and long-distance circulation. Such lanceolate
blades, although smaller, continue into the early seventeenth century
when they serve as regalia among political and social elites in ritual
performances.
